---
title: 'Title'
author: 'Florian Muthreich'
date: '2020/02/04'
output: pdf_document
---


Notes:
- Time to measure 50 pollen grains on single taxa slides: 60 min. 
- includes Mounting, selecting pollen, measuring time FTIR

Fossil pollen will require more searching in the selection phase.
Have to measure 75 grains if we target 50 good spectra.
Estimate time for mounting, selecting, measuring 75 grains of one taxa: 90 - 120 min

10 days in the lab; 8+ hrs a day: 80 hrs
Target: 5 taxa per sample; 5 x 90: 7.5 hrs `r 5*90/60`

9-12 13-24
10 days in the lab; 14 hrs a day: 140
Target: 5 taxa per sample; 140/7.5 = 18; 14/10 = 14

```{r}
total_samples = function(hrs_per_sample, no_taxa, total_hrs = 140){
  tmp = total_hrs / (no_taxa * hrs_per_sample)
  return(signif(tmp,3))
}
M = outer(c(1,1.5,2),c(3,4,5),total_samples, total_hrs = 140)
row.names(M) = c(1,1.5,2)
colnames(M) = c(3,4,5)
M
```

Anywhere from 18 to 30 samples is reasonable.
At 3 cores 5 taxa:
Minimum 4 samples per core 
Maximum 9 samples per core
```{r}
signif(M/3, 3)
```
At 2 cores 5 taxa:
Minimum 7 samples per core
Maximum 14 samples per core
```{r}
signif(M/2, 3)
```

