---
title: 'Chemistry of fossil Pinus pollen and the effect of acetolysis'
date: '2020/02/04'
output:
  bookdown::pdf_document2:
    toc: false
    extra_dependencies: ["flafter"]
bibliography: references.bib
csl: biology-letters.csl
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here("R","packages.R"))
source(here("R","authors_and_affiliations.R"))
```

---
author: `r Authors`
---
`r Affiliations`

2500 words not including in tables title page references
2 figs and 2 tables
figures separate legends in text
separate cover letter

<!-- 
A  200
I  500
MM 400
R  500
D  800
-->

\newpage

# Abstract
200 words no abreviations/refs

Keywords: FTIR, fossil pollen, Holocene, density separation, acetolysis, pollen chemistry

\newpage
<!-- READ DATA -->
```{r message=FALSE, warning=FALSE, include=FALSE}
source(here("R","ages.R"))
if (file.exists(here("data","output","DAL_spectra.rds"))) {
  DAL = readRDS(here("data","output","DAL_spectra.rds"))
  DAL$age = 0
} else {
  source(here("R","read_data.R"))
  read_data("DAL")
  DAL = readRDS(here("data","output","DAL_spectra.rds"))
}

if (file.exists(here("data","output","MFM_spectra.rds"))) {
  MFM = readRDS(here("data","output","MFM_spectra.rds"))
  MFM[MFM$depth=="A4UR_72-74", 'depth'] = "A5UR_72-74" #fixing typo in sample name
  MFM$age = fn_age(MFM$depth)
  MFM[MFM$treatment == "","treatment"] = "SPT"
} else {
  source(here("R","read_data.R"))
  read_data("MFM")
  MFM = readRDS(here("data","output","MFM_spectra.rds"))
}

if (file.exists(here("data","output","TSK_spectra.rds"))) {
  TSK = readRDS(here("data","output","TSK_spectra.rds"))
  TSK$age = fn_age(TSK$depth)
} else {
  source(here("R","read_data.R"))
  read_data("TSK")
  TSK = readRDS(here("data","output","TSK_spectra.rds"))
}

if (file.exists(here("data","output","fresh_spectra.rds"))) {
  FRE = readRDS(here("data","output","FRE_spectra.rds"))
} else {
  source(here("R","read_data.R"))
  read_data_fresh()
  FRE = readRDS(here("data","output","FRE_spectra.rds"))
}

FRE$Label = "FRE"
DAL$Label = "DAL"
TSK$Label = "TSK"
MFM$Label = "MFM"

outliers = c("DAL_rx1GZ", "DAL_2cxp7", "DAL_VzFi0", "DAL_ANdP0", "MFM_av67v", "MFM_OftM2", "MFM_xogJr", "MFM_hoG4y", "MFM_5UNIJ", "TSK_vvgOZ", "DAL_xgYjM", "TSK_m4hN6", "TSK_e04Go") #THese MFM ids dont exist anymore

DAL = DAL[!DAL$ids %in% outliers,]
TSK = TSK[!TSK$ids %in% outliers,]
MFM = MFM[!MFM$ids %in% outliers,]
```

<!-- PRE-PROCESS DATA -->
```{r}
if (!hasName(DAL, "emsc")) {
  source(here("R","pre_process.R"))
}
```

<!-- PCA -->
```{r}

```


```{r DALcounts, echo=FALSE, results='asis'}
count_grains = function(core){
  counts = core %>%
    subset(type == "pollen" & species == "Pinus") %>% 
    group_by(type, depth, treatment) %>%
    summarise(count=n())
  return(counts)
}
count_grains = function(core){
  counts = core %>%
    subset(type == "pollen") %>% 
    group_by(treatment, depth, species) %>%
    summarise(count=n())
  return(counts)
}
kable(count_grains(DAL), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Dalmutladdo core.")
#      digits = 1, format.args = list(scientific = F),
#      col.names = c("")) %>% 
#  add_header_above()
```

```{r TSKcounts, echo=FALSE, results='asis'}
kable(count_grains(TSK), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Tiefer See core.")
```

```{r MFMcounts, echo=FALSE, results='asis'}
kable(count_grains(MFM), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Meerfelder Maar core.")
```

\newpage

# Introduction
<!--Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.-->

The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of palaeoecology and palaeobotany.
Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and are important tools to address ecological processes with timescales that operate on centennial/decadal timescales.

The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
The structure of sporopollenin has been described as similar to lignins, e.g. importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], but the exact structure remains unknown.

In recent years, the study of pollen-grain chemistry has shown promising results for biological and palaeoecological research.
Fourier Transform Infrared (FTIR) spectroscopy has become a popular method  for characterising the chemical variation within pollen and spores.
A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020]. 
This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution @dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016.
Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020].
Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes.

So far, most FTIR studies have used modern material or material extracted from herbarium specimens .
In some cases the pollen samples are processed with acetolysis following standard palynological methods [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018].
The assumption is that acetolysis does not alter the chemistry of sporopollenin and is not affecting any chemical signal.
To date, there has not been a comparison between acetolysed and untreated fossil pollen to examine any effect of acetolysis on sporopollenin or chemical degradation of pollen during diagenesis.

In this study we investigate the chemistry of fossil pollen to explore three potential sources of chemical alteration: i. the effects of diagenesis, ii. acetolysis and iii. age on pollen chemistry.
We extracted pollen from three sediment cores using density separation to obtain chemically untreated fossil spectra.
The pollen was collected from three different cores in three locations (Germany and Norway), which allows us to examine potential differences in diagenesis on pollen chemistry at different locations.
One replicate of the pollen samples in each core were acetolysed to analyse any changes caused by the strong acid treatment and compared to the untreated replicate.
Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.

RQs
i. Differences between cores
ii.i Difference between acetolysed and fossil
ii.ii Difference between fresh and fossil grains.
iii effect of depth/age on chemistry

iv effect of orientation on chemistry

    
acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility

# Methods and Materials

The pollen was extracted from three cores: Dalmuttlado in Norway, Tiefer See [@dragerVarveMicrofaciesVarve2017] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
Tiefer See formed during the last glacial period, while the Meerfelder Maar record is from the largest maar of the Westeifel Volcaniv field and formed ~35000 BP.
Tiefer See and Meerfelder Maar are varved cores and were the samples used in this study span two different time-periods.
The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).  
The samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) at 1.43 (g/ml)? as separation medium. <!--FM: check Unit-->
During density separation a known volume of SPT (density 1.43) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
After centrifugation, particulates/material heavier than 1.43 has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
In our case the lighter fraction contained most of the pollen.
The density of pollen is reported at <1.4 UNIT [@nakagawaDensemediaSeparationMore1998; @regnellPreparativeCentrifugationNew1996].

Preprocessing:
- SG 2nd degree, 2 poly, windowsize 5 or 6
- EMSC linear (probably needs to involve background spectra to filter them out)

PCA on grains for exploration

PLS regression of spectra on depth

repeat on different grain averages (1, 3, 5, 10)

average all spectra and stack plots of different depth

Pollen extracted from sediment samples 
[@kampsAgentSeparationDissolved1985] SPT patent
[@nakagawaDensemediaSeparationMore1998] 
[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4

\newpage

# Results

i. Differences between cores
ii.i Difference between acetolysed and fossil
ii.ii Difference between fresh and fossil grains.
iii effect of depth/age on chemistry

iv effect of orientation on chemistry


we see a peak in the SPT wide spectra at 1090 - 1100 that is not caused by pollen.
wide peak indicates inorganic compound.
This peak is not present in either fresh pollen or acetolysed pollen.
1090 peak is present in all three cores, in addition Tiefer See shows a similar lower peak at 1620 - 1630.
The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170 in all three cores.
PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into cores (Fig pca_acet_sel)

There are differences between fresh and fossil grains.
Most notibly the lipid peak at 1745 is not visible in the fossil grains, neither acetolysed or raw fossil pollen.
Changes in the carbohydrate region from 1000 to 1250.
in SPT pollen there is a dominant peak there.
The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.


1170 is more pronounced in spt than acet.
the 1090 peak dissappear after acetolysis from all spectra as does the 1630 in TSK. 
a peak at 1040 is now visible in acetolysed pollen that was covered by the wide peak in the SPT pollen.
The region 1000 - 1200 is now dominated by two peaks 1040 and 1170 and its shoulders.
In the PCA ordination the acetolysed pollen is separated from the SPT pollen, stronger separation for TSK than for DAL and MFM.
A LDA on each core separately had `r signif(mean(lda_tables[[1]]$acc),1)` % accuracy to distinguish between sample treatment (Table LDA treatment).

LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen.
For the SPT pollen the accuracy increased when the depths were restricted to the same depths as the acetolysed samples.
Generally lower number of depths to classify increased accuracy.
sporopollenin peaks are similar between depths, changes in the other parts of the spectra are driving separation. 



```{r message=FALSE, warning=FALSE, include=FALSE}
pca_acet_select = readRDS(here("data","output","pca_plots.rds"))
lda_tables = readRDS(here("data","output","lda_tables.rds"))
```

```{r pca_acet_sel echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
pca_acet_select[1]
```

## Spectra mean plots
```{r echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
mean_plots = readRDS(here("data","output","mean_plots.rds"))

mean_plots[[7]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
mean_plots[[1]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
mean_plots[[2]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
mean_plots[[3]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
mean_plots[[4]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
mean_plots[[5]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
mean_plots[[6]]
```

\newpage

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
blank_mean_plots[[1]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
blank_mean_plots[[2]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
blank_mean_plots[[3]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
blank_mean_plots[[4]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
blank_mean_plots[[5]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
blank_mean_plots[[6]]
```


# Discussion

strange peak in SPT samples most likely biogenic silica.
is removed with acetolysis (check references of this happening anywhere else)

some effect of the core on chemistry, because TSK is more dissimilar than the other two cores and TSK is laminated sediment.
BSi peak at 1630 was only present here.
TSK samples are also youngest, most dissimilar between SPT and acetolysed pollen, maybe because these pollen had still remnants of non sporopollenin parts that were removed by acetolysis

acetolysation had an effect on pollen chemistry.
especially in the 1000-1200 region, which had the BSi peak but also the region from 1150 to 1250 changes, which is outside of the main effects of the BSi peak. These regions are mainly caused by carbohydrates, which could be cellulose from the grain wall.

the pollen spectra are way different from fresh pollen.
the sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g. carbohydrates).
GCMS on these pollen can give more answers

no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.


Paragraph on practicality
This is an improvement on previous single grain measurements with hyperion, where each grain had to be targeted individually.
instead a large area could be measured and spectra selected post-hoc.
This introduced also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
Nevertheless this introduces another challenge: concentration of pollen.
Some of our samples had very low concentration of pollen

# References
