---
title: 'Viability of FTIR on fossil pollen'
author: 'Florian Muthreich'
date: '2020/02/04'
output:
  pdf_document: default
bibliography: references.bib
---
# Notes
extracted fossil pollen from sediment.
ca 5 common taxa: Alnus, Picea, Pinus,

## Cores
Dalmuttlado:
- pine
- holocene

Totenmoos [@heissUltenValleySouth2005]:
- bog-sediment core; past 15000 yrs
- min 1000 grains counted
- Picea, Pinus very common, switch at 10000 BP from Pinus to Picea
- Alnus, Betula throughout
- Abies, Fagus after 9000 BP
- trace Quercus

Younger Dryas [@brauerHighResolutionSediment1999]:
- 12 samples laminated sediment
- from  11400 to 13120
- min 500 grains
- pollen conc highest in 11000-11600 and 12700-12900
- Pinus betula most common (ca 150 grains counted)
- salix (ca 25 counted)
- quercus throughout in low conc (<1%)
- graminoids/cyperaceae/artemisia common

Tiefer See:
- 1870-2010
- laminated sediments
- Pinus pollen

Tinner alpine:

Tinner meso-mediterranean:
[@bisculmHoloceneVegetationFire2012; @vanniereClimateHumandrivenFire2008]
Quercus sections Quercus, Cerris and Ilex

Tinner sub-mediterranean:
[@zoller19601960]
Quercus sections Quercus, Cerris and Ilex

## Time to measure notes:
- Time to measure 50 pollen grains on single taxa slides: min 60 min. 
- includes Mounting, selecting pollen, measuring time FTIR

Fossil pollen will require more searching in the selection phase.
Have to measure ~75 grains if we target 50 good spectra.
Estimate time for mounting, selecting, measuring 75 grains of one taxa: 90 - 120 min

10 days in the lab; 8+ hrs a day: 80 hrs
Target: 5 taxa per sample; 5 x 90: 7.5 hrs `r 5*90/60`

9-12 13-24
10 days in the lab; 14 hrs a day: 140
Target: 5 taxa per sample; 140/7.5 = 18; 14/10 = 14

```{r}
total_samples = function(hrs_per_sample, no_taxa, total_hrs = 140){
  tmp = total_hrs / (no_taxa * hrs_per_sample)
  return(signif(tmp,3))
}
M = outer(c(1,1.5,2),c(3,4,5),total_samples, total_hrs = 140)
row.names(M) = c(1,1.5,2)
colnames(M) = c(3,4,5)
print('Number of samples at One Core')
M
```

Anywhere from 18 to 30 samples is reasonable.
At 3 cores 5 taxa:
Minimum 4 samples per core 
Maximum 9 samples per core

```{r}
print('No. of samples at Three Cores')
signif(M/3, 3)
```
At 2 cores 5 taxa:
Minimum 7 samples per core
Maximum 14 samples per core

```{r}
print('No. of samples at Two Cores')
signif(M/2, 3)
```

# Introduction
Research Questions:
1. chemical variation between taxa in the same cores.
2. chemical variation between taxa between cores 
3. chemical variation of pollen with depths/age

# Methods and Materials
## Labwork Report

did µFTIR on fresh and fossil pollen using Hyperion-FPA

This is very sample dependent.
organic rich cores are problematic (requires a better extraction protocol), pollen density is another.
Dalmutladdo was a very good core for this: low in organic and relatively rich in pollen.
BSi increased with depth, highest in oldest samples (extracted blanks).
Still, some depth were more time intensive (130/150), low amount of pollen .
I managed about 2-3 core samples per day, the fresh samples can be done in a fraction of the time, all done within 1 half day.

9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (see table)

possibilities for Alnus and Betula, even though scattering visible in spectra 

```{r}
dalmutladdo = data.frame(ID = c("DAL_08","DAL_20","DAL_40","DAL_60","DAL_90", "DAL_110","DAL_130","DAL_140","DAL_150"),
                         pinus_pollen = c(23,49,37,47,36,32,39,17,7),
                         alnus_pollen = c(0,7,0,0,0,0,0,0,0),
                         betula_pollen = c(0,15,0,0,0,0,0,0,0))
dalmutladdo
```

```{r}
fresh_pollen = data.frame(ID = c("BGO_PS7", "BGO_PS8", "INN_PS13", "INN_PS19A","NMBU_Pin_syl_14E1", "NMBU_Pin_syl_14H1", "NMBU_Aln_glu_14E1", "NMBU_Aln_glu_14J1", "NMBU_Aln_inc_14B1", "NMBU_Aln_inc_14G1", "NMBU_Bet_pub_14A1", "NMBU_Bet_pub_14B1", "NMBU_Cor_ave_14B1", "NMBU_Cor_ave_14C1"),
                          extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
fresh_pollen
```


Research questions we could answer with this set:

- Differences between fresh and fossil/near fossil grains of Pinus
- chemical changes in Pinus pollen with age/depth
(- chemical differences between Pinus/Alnus/Betula)

extensions to this set:
- add more samples from Dalmuttladdo highest and lowest depth (maybe labwork)
- add cores (e.g. MFM, Tiefer See or Italian cores), Festi core was quite organic rich
- 

# Introduction

# Methods and Materials 
PCA on grains for exploration

regression of spectra on depth

repeat on different grain averages (1, 3, 5, 10)

average all spectra and stack plot

Pollen extracted from sediment samples 
[@kampsAgentSeparationDissolved1985] SPT patent
[@nakagawaDensemediaSeparationMore1998] 
[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4
