---
title: 'Chemistry of fossil Pinus pollen and the effect of acetolysis'
date: '2020/02/04'
output:
  bookdown::pdf_document2:
    toc: false
    extra_dependencies: ["flafter"]
bibliography: references.bib
csl: biology-letters.csl
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here("R","packages.R"))
source(here("R","authors_and_affiliations.R"))
```

---
author: `r Authors`
---
`r Affiliations`

2500 words not including in tables title page references
2 figs and 2 tables
figures separate legends in text
separate cover letter

<!-- 
A  200
I  500
MM 400
R  500
D  800
-->

\newpage

# Abstract
200 words no abreviations/refs

Keywords: FTIR, fossil pollen, Holocene, density separation, acetolysis, pollen chemistry

\newpage
<!-- READ DATA -->
```{r message=FALSE, warning=FALSE, include=FALSE}
source(here("R","ages.R"))
if (file.exists(here("data","output","DAL_spectra.rds"))) {
  DAL = readRDS(here("data","output","DAL_spectra.rds"))
  DAL$age = 0
} else {
  source(here("R","read_data.R"))
  read_data("DAL")
  DAL = readRDS(here("data","output","DAL_spectra.rds"))
}

if (file.exists(here("data","output","MFM_spectra.rds"))) {
  MFM = readRDS(here("data","output","MFM_spectra.rds"))
  MFM[MFM$depth=="A4UR_72-74", 'depth'] = "A5UR_72-74" #fixing typo in sample name
  MFM$age = fn_age(MFM$depth)
  MFM[MFM$treatment == "","treatment"] = "SPT"
} else {
  source(here("R","read_data.R"))
  read_data("MFM")
  MFM = readRDS(here("data","output","MFM_spectra.rds"))
}

if (file.exists(here("data","output","TSK_spectra.rds"))) {
  TSK = readRDS(here("data","output","TSK_spectra.rds"))
  TSK$age = fn_age(TSK$depth)
} else {
  source(here("R","read_data.R"))
  read_data("TSK")
  TSK = readRDS(here("data","output","TSK_spectra.rds"))
}

if (file.exists(here("data","output","fresh_spectra.rds"))) {
  FRE = readRDS(here("data","output","FRE_spectra.rds"))
} else {
  source(here("R","read_data.R"))
  read_data_fresh()
  FRE = readRDS(here("data","output","FRE_spectra.rds"))
}

FRE$Label = "FRE"
DAL$Label = "DAL"
TSK$Label = "TSK"
MFM$Label = "MFM"

outliers = c("DAL_rx1GZ", "DAL_2cxp7", "DAL_VzFi0", "DAL_ANdP0", "MFM_av67v", "MFM_OftM2", "MFM_xogJr", "MFM_hoG4y", "MFM_5UNIJ", "TSK_vvgOZ", "DAL_xgYjM", "TSK_m4hN6", "TSK_e04Go") #THese MFM ids dont exist anymore

DAL = DAL[!DAL$ids %in% outliers,]
TSK = TSK[!TSK$ids %in% outliers,]
MFM = MFM[!MFM$ids %in% outliers,]
```

<!-- PRE-PROCESS DATA -->
```{r}
if (!hasName(DAL, "emsc")) {
  source(here("R","pre_process.R"))
}
```

<!-- PCA -->
```{r}

```


```{r DALcounts, echo=FALSE, results='asis'}
count_grains = function(core){
  counts = core %>%
    subset(type == "pollen" & species == "Pinus") %>% 
    group_by(type, depth, treatment) %>%
    summarise(count=n())
  return(counts)
}
count_grains = function(core){
  counts = core %>%
    subset(type == "pollen") %>% 
    group_by(treatment, depth, species) %>%
    summarise(count=n())
  return(counts)
}
kable(count_grains(DAL), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Dalmutladdo core.")
#      digits = 1, format.args = list(scientific = F),
#      col.names = c("")) %>% 
#  add_header_above()
```

```{r TSKcounts, echo=FALSE, results='asis'}
kable(count_grains(TSK), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Tiefer See core.")
```

```{r MFMcounts, echo=FALSE, results='asis'}
kable(count_grains(MFM), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Meerfelder Maar core.")
```

\newpage

# Introduction
<!--Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.-->

The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of palaeoecology and palaeobotany.
Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and are important tools to address ecological processes with timescales that operate on centennial/decadal timescales.

The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
The structure of sporopollenin has been described as similar to lignins, e.g. importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], but the exact structure remains unknown.

In recent years, the study of pollen-grain chemistry has shown promising results for biological and palaeoecological research.
Fourier Transform Infrared (FTIR) spectroscopy has become a popular method  for characterising the chemical variation within pollen and spores.
A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020]. 
This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution @dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016.
<!--MOVE TO DISCUSSION: Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020].
Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes. -->


So far, most FTIR studies focused on identification of pollen have used modern material and bulk measurements, while some have used FTIR microscopes.
These FTIR microscopes required manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
FTIR as a method has advanced on a technical level over the past 5 to 10 years.
The development and capabilities of microscopes with connected FTIR spectrometers now allow easier and faster capture of spectra of pollen grains.
With these newer machines a larger area is irradiated and spectra are recorded at higher resolution over the entire irradiated area.
This allows capture of an image and matrix of spectra over the same sample area, which enables e.g. the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide.

Palaeoecologists use a combination of several chemical extractions to isolate and concentrate pollen from the sediment matrix.
Most common is acetolysis in addition to HF treatment and HCl [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
These strong acid treatments aim to remove other organic particulates, calcium carbonates and silicates, while preserving morphological integrity of pollen grains [REF].
A limited number of studies have used acetolysis and other chemical treatments in combination with FTIR on pollen to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;@schulzeosthoffPhenolsIntegratedCompounds1987;@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
These results suggest that acetolysis may alter the chemical composition of sporopollenin.

An alternative to chemical extraction is density separation using e.g. sodium polytungstate (SPT), which is used to isolate pollen from minerogenic sediments [REFS] or as preparations for pollen isotope studies [REFS].
Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of sporomorph chemistry.
SPT is used to adjust the density of sample suspension and facilitace separation of less dense particulates (e.g. pollen and other microfossils) over mineral particles.
SPT is removed from the sample with several washes after extraction and any contamination of the sample is easily identifiable by characteristic wavebands in the spectra. 

In this study we explore the use of SPT density separation as an alternative to acetolysis by investigating the chemical alteration of SPT and acetolyis on modern pollen and apply both extraction methods to three fossil cores.
With this, we aim to investigate changes in fossil pollen chemistry, due to: i. differences between modern and fossil material, ii. extraction method and iii. effect of diagenesis (age of fossil pollen).
The effect of treatment with SPT on pollen chemistry is tested on modern material to establish if any chemical changes occur.
The pollen was collected from three different cores in three locations (Germany and Norway), which allows us to examine potential differences in diagenesis on pollen chemistry at different locations.
Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
The pollen was extracted from the sediment cores using density separation, with one replicate being acetolysed afterwards.


RQs
i. Differences between cores
ii.i Difference between acetolysed and fossil
ii.ii Difference between fresh and fossil grains.
iii effect of depth/age on chemistry

iv effect of orientation on chemistry


# Methods and Materials

The pollen was extracted from three cores: Dalmuttlado in Norway, Tiefer See [@dragerVarveMicrofaciesVarve2017] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
Tiefer See formed during the last glacial period, while the Meerfelder Maar record is from the largest maar of the Westeifel Volcaniv field and formed ~35000 BP.
Tiefer See and Meerfelder Maar are varved cores and were the samples used in this study span two different time-periods.
The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).  
The samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
During density separation a known volume of SPT (density 1.43 g cm^-3^) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
After centrifugation, particulates/material heavier than 1.43 g cm^-3^ has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
In our case the lighter fraction contained most of the pollen.
There is no agreement on the density of the pollen exine and reported values vary between  <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].

## FTIR measurements

The spectra were measured with a Vertex 70 FTIR spectrometer connected to a Hyperion 3000 IR microscope (Bruker Optik, Ettlingen, Germany) equipped with a focal plane array (FPA) nitrogen cooled mercury cadmium telluride (MCT) detector and a 15x magnification objective.
The FPA-MCT detector allows the capture of FTIR images, where each pixel is a full spectra, instead point measurements.
For this study images of 128x128 pixels were taken and each spectra was recorded with 64 scans.
The accompanied visible images were obtained by a charge coupled device (CCD) camera coupled to the microscope.
Single grain pollen grain spectra were extracted from the FTIR images, by averaging all pixel spectra in the image area of the pollen grain. 
Data acquisition and instrument control was performed using the OPUS 7.5 software (Bruker Optik GmbH, Germany).

The extracted pollen was applied to ZnSe slides and a background spectra was measured before each sample on an empty area of the slide.
The pollen treated with SPT and acetolysis were applied onto the slide suspended in water and dried out before measurement under a fume hood.
The extracted sediment samples contained other small material than pollen (e.g. diatom fragments, organic material fragments, etc).
We recorded several "blank" spectra of dried residue that formed over the dried pollen residue (see also SM).

## Spectra preprocessing and analysis 

Preprocessing:
- SG 2nd degree, 2 poly, windowsize 5 or 6
- EMSC linear (probably needs to involve background spectra to filter them out)

PLSR second deriv on treatment (fresh vs fossil)

PLS regression of spectra on depth

repeat on different grain averages (1, 3, 5, 10)

average all spectra and stack plots of different depth



\newpage

```{r message=FALSE, warning=FALSE, include=FALSE}
pca_acet_select = readRDS(here("data","output","pca_plots.rds"))
lda_tables = readRDS(here("data","output","lda_tables.rds"))
```

# Results

i. Differences between cores
ii.i Difference between acetolysed and fossil
ii.ii Difference between fresh and fossil grains.
iii effect of depth/age on chemistry


There are differences between modern and fossil grains (Fig. \@ref(fig:meanDAL), \@ref(fig:meanMFM) and \@ref(fig:meanTSK).
The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
Most notibly the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra, while we see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.

After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:plsrord)).
The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:plsrcoef)).



<!-- 
Tiefer See pollen is separated from the other two cores, which cluster closer together.
This is the same for acetolysed and SPT pollen. 
A PCA-LDA model to differentiate fossil pollen into the three cores achieved ~60% accuracy to classify MFM and DAL pollen and 85% accuracy to identify TSK pollen (REF TO SM). 

-->


the peak at 1170 cm^-1^ is more pronounced in spt than acetolyzed pollen.

the 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in TSK. 
a peak at 1040 cm^-1^ is now visible in acetolysed pollen that is not visible in the SPT pollen.
The region 1000 - 1200 cm^-1^ is now dominated by two peaks at 1040 and 1170 and its shoulders.
In the PCA ordination the acetolysed pollen is separated from the SPT pollen, stronger separation for TSK than for DAL and MFM.
A LDA on each core separately had `r signif(mean(lda_tables[[1]]$acc),1)` % accuracy to distinguish between sample treatment (Table \@ref(tab:ldatreatment)).

LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen (Table \@ref(tab:ldadepth)).
For the SPT pollen the accuracy increased when the depths were restricted to the same depths as the acetolysed samples.
Generally lower number of depths to classify increased accuracy.
sporopollenin peaks are similar between depths, changes in the other parts of the spectra are driving separation.


```{r pcaacetsel, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
pca_acet_select[1]
```

```{r ldatreatment, echo=FALSE, results='asis'}

kable(lda_tables[["treatment"]], booktabs = T, format = "latex",
      caption = "Accuracy of bootstrapped PCA LDA to distinguish treatment (SPT vs acet)")
```

```{r ldadepth, echo=FALSE, results='asis'}

kable(lda_tables[["depth"]], booktabs = T, format = "latex",
      caption = "Accuracy of bootstrapped PCA LDA to distinguish depths for each core. treatments are separate.")
```

## Spectra mean plots

```{r meanFRE, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))

mean_plotsSM[[7]]
```

```{r meanDAL, echo=FALSE, fig.cap="DAL whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))

mean_plots[[1]]
```

```{r meanMFM, echo=FALSE, fig.cap="MFM whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
mean_plots[[2]]
```

```{r meanTSK, echo=FALSE, fig.cap="TSK whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
mean_plots[[3]]
```


# Discussion

.Overall intro to discussion

.difference to fresh pollen
the fossil pollen spectra show changes from fresh pollen.
lipid and protein peaks have dissappeared and there is a lot of differences in the carbohydrate region.
this part of the spectra has changed substantially and most fresh components are most likely removed.
Nevertheless some peaks in the carb region indicate some amount of cellulose or grain material still present.
the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g. carbohydrates).

.on bsi peak in spt samples
wide peak indicates inorganic compound.
This peak in the SPT samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016].
we observed a crystalline layer over our pollen grains, when observed under the microscope. 
this peak is removed with acetolysis (check references of this happening anywhere else)

.acetolysis effects
acetolysation had an effect on pollen chemistry.
especially in the 1000-1200 region, which had the BSi peak but also the region from 1150 to 1250 changes, which is outside of the main effects of the BSi peak. These regions are mainly caused by carbohydrates, which could be cellulose from the grain wall.

GCMS on these pollen can give more answers

.core effects
some effect of the core on chemistry, because TSK is more dissimilar than the other two core, could be connected to the fact that TSK is laminated sediment.
BSi peak at 1630 was only present here.

We saw large overlap between MFM and DAL in the PCA ordination and lower accuracies in LDA differentiating between cores, both in SPT and acetolysed pollen.
This seems to indicate that pine pollen from different cores are chemically quite similar, even cores that are far apart.
TSK pollen on the other hand could be clearly separated from the other two cores, in both SPT and acetolysed pollen. 
In fact acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than SPT pollen is.
There seem to be chemical differences between TSK and the other two cores that are present in SPT and acetolysed pollen.
One reason for the dissimilarity of SPT TSK pollen is the BSi peak at 1630 which is only present in TSK, but even acetolysed pollen from TSK is quite different from the other cores according to PCA and LDA analysis.
The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene.
The other difference is that TSK is a varved core, while MFM and DAL are not.
Our findings suggest a chemical difference in fossil pollen grains, either caused by diagenesis (varved vs non-varved), age or combination of the two that is persistent even after acetolysis. 


.depth patterns
no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.

.application for palaeoecology
Paragraph on practicality
This is an improvement on previous single grain measurements with hyperion, where each grain had to be targeted individually.
instead a large area could be measured and spectra selected post-hoc.
This introduces also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
Nevertheless this introduces another challenge: concentration of pollen.
Some of our samples had very low concentration of pollen and would need a concentration step in sample processing.
Our results do show that acetolysis does not remove any of the sporopollenin peaks, but some of the other parts of the spectra, and could therefore be seen as a standardizing procedure.
Further studies with GCMS methods on fossil pollen and several species have to be performed. 

.conclusions

# Mean spectra figures SM


```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
mean_plotsSM[[1]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
mean_plotsSM[[2]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
mean_plotsSM[[3]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
mean_plotsSM[[4]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
mean_plotsSM[[5]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
mean_plotsSM[[6]]
```

\newpage

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
blank_mean_plots[[1]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
blank_mean_plots[[2]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
blank_mean_plots[[3]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
blank_mean_plots[[4]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
blank_mean_plots[[5]]
```

```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
blank_mean_plots[[6]]
```



# References
