---
title: 'Viability of FTIR on fossil pollen'
author: 'Florian Muthreich'
date: '2020/02/04'
output: pdf_document
---

extracted fossil pollen from sediment.
ca 5 common taxa: Alnus, Picea, Pinus, 

Research Questions:
1. chemical variation between taxa in the same cores.
2. chemical variation between taxa between cores 
3. chemical variation of pollen with depths/age

Notes:
- Time to measure 50 pollen grains on single taxa slides: min 60 min. 
- includes Mounting, selecting pollen, measuring time FTIR

Fossil pollen will require more searching in the selection phase.
Have to measure ~75 grains if we target 50 good spectra.
Estimate time for mounting, selecting, measuring 75 grains of one taxa: 90 - 120 min

10 days in the lab; 8+ hrs a day: 80 hrs
Target: 5 taxa per sample; 5 x 90: 7.5 hrs `r 5*90/60`

9-12 13-24
10 days in the lab; 14 hrs a day: 140
Target: 5 taxa per sample; 140/7.5 = 18; 14/10 = 14

```{r}
total_samples = function(hrs_per_sample, no_taxa, total_hrs = 140){
  tmp = total_hrs / (no_taxa * hrs_per_sample)
  return(signif(tmp,3))
}
M = outer(c(1,1.5,2),c(3,4,5),total_samples, total_hrs = 140)
row.names(M) = c(1,1.5,2)
colnames(M) = c(3,4,5)
print('Number of samples at One Core')
M
```

Anywhere from 18 to 30 samples is reasonable.
At 3 cores 5 taxa:
Minimum 4 samples per core 
Maximum 9 samples per core
```{r}
print('No. of samples at Three Cores')
signif(M/3, 3)
```
At 2 cores 5 taxa:
Minimum 7 samples per core
Maximum 14 samples per core
```{r}
print('No. of samples at Two Cores')
signif(M/2, 3)
```

