---
title: 'Chemistry of fossil Pinus pollen and the effect of acetolysis'
date: '2020/02/04'
output:
  bookdown::pdf_document2:
    toc: false
    extra_dependencies: ["flafter"]
bibliography: references.bib
csl: biology-letters.csl
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here("R","packages.R"))
source(here("R","authors_and_affiliations.R"))
```

---
author: `r Authors`
---
`r Affiliations`

2500 words not including in tables title page references
2 figs and 2 tables
figures separate legends in text
separate cover letter

\newpage

# Abstract
200 words no abreviations/refs

Keywords: FTIR, fossil pollen, Holocene, density separation, acetolysis, pollen chemistry

\newpage
<!-- READ DATA -->
```{r}
if (file.exists(here("data","output","DAL_spectra.rds"))) {
  DAL = readRDS(here("data","output","DAL_spectra.rds"))
} else {
  source(here("R","read_data.R"))
  read_data("DAL")
  DAL = readRDS(here("data","output","DAL_spectra.rds"))
}

if (file.exists(here("data","output","MFM_spectra.rds"))) {
  MFM = readRDS(here("data","output","MFM_spectra.rds"))
} else {
  source(here("R","read_data.R"))
  read_data("MFM")
  MFM = readRDS(here("data","output","MFM_spectra.rds"))
}

if (file.exists(here("data","output","TSK_spectra.rds"))) {
  TSK = readRDS(here("data","output","TSK_spectra.rds"))
} else {
  source(here("R","read_data.R"))
  read_data("TSK")
  TSK = readRDS(here("data","output","TSK_spectra.rds"))
}
```

<!-- PRE-PROCESS DATA -->
```{r}
if (!hasName(DAL, "sg2")) {
  source(here("R","pre_process.R"))
}
```

<!-- PCA -->
```{r}
DAL$Label = "DAL"
TSK$Label = "TSK"
MFM$Label = "MFM"

pca_data = rbind(subset(DAL, species == "Pinus"), 
        subset(TSK, species == "Pinus"), 
        subset(MFM, species == "Pinus"))
pca_data[pca_data$treatment == "","treatment"] = "SPT"
pca_res = princomp(pca_data$sg2, scores = T)
scores =  as.data.frame(pca_res$scores[,1:10]) %>%
  mutate(ids = rownames(.), 
         core = pca_data$Label, 
         treatment = pca_data$treatment
         ) #%>%
  #melt(.,id.vars = "ids")
ggplot(scores, aes(Comp.1, Comp.2, color = core)) + geom_point()
# pca_data[scores[scores[,2] < -.2,"ids"]==pca_data$ids,1:10]
```


```{r DALcounts, echo=FALSE, results='asis'}
count_grains = function(core){
  counts = core %>%
    subset(type == "pollen" & species == "Pinus") %>% 
    group_by(type, depth, treatment) %>%
    summarise(count=n())
  return(counts)
}
kable(count_grains(DAL), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Dalmutladdo core.")
#      digits = 1, format.args = list(scientific = F),
#      col.names = c("")) %>% 
#  add_header_above()
```

```{r TSKcounts, echo=FALSE, results='asis'}
kable(count_grains(TSK), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Tiefer See core.")
```

```{r MFMcounts, echo=FALSE, results='asis'}
kable(count_grains(MFM), booktabs = T, format = "latex",
      caption = "Pinus pollen grain count in Meerfelder Maar core.")
```

\newpage

# Introduction
<!--Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.-->

The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of Palaeoecology and Palaeobotany.
Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @nogues-bravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013].

The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
The structure of sporopollenin has been described as similar to lignins, e.g. importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], but the exact structure remains unknown.

In recent years, the study of pollen grain chemistry has shown promising results for biological and palaeoecological research.
Fourier Transform Infrares (FTIR) spectroscopy has become a popular method for analysing pollen chemistry.
A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020]. 
This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution @dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016.
Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020].
Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes.

So far, most FTIR studies have used modern material or material extracted from herbarium specimens .
In some cases the pollen samples are processed with acetolysis following standard palynological methods [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018].
The assumption is that acetolysis does not alter the chemistry of sporopollenin and is not affecting any chemical signal.
To date, there has not been a comparison between acetolysed and untreated fossil pollen to examine any effect of acetolysis on sporopollenin or chemical degradation of pollen during diagenesis.

In this study we investigate the chemistry of fossil pollen to explore three potential sources of chemical alteration: i. the effects of diagenesis, ii. acetolysis and iii. age on pollen chemistry.
We extracted pollen from three sediment cores using density separation to obtain chemically untreated fossil spectra.
The pollen was collected from three different cores in three locations (Germany and Norway), which allows us examine potential differences in diagenesis on pollen chemistry at different locations.
One replicate of the pollen samples in each core were acetolysed to analyse any changes caused by the strong acid treatment and compared to the untreated replicate.
Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.


RQs
i. Difference between fresh and fossil grains.
i.i effect of orientation on chemistry
i.ii effect of depth/age on chemistry
ii Differences between cores
    
acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility

# Methods and Materials

The pollen was extracted from three cores: Dalmuttlado in Norwa

Three cores from different environments and age ranges

Pollen extraction:

1. add 1 cube to vial (~1cm^-3^)
2. wash with water -> dissolve sediment vortex
3. centrifuge 3min 3000rpm
4. decant; save decant
5. add 3 ml 1.43 SPT solution
6. vortex
7. centrifuge 3min 3000rpm
8. decant supernatant into new vial
9. repeat 5-7; decant into same vial
10. wash with water; save decant for checking
11. wash with water; save decant x2
12. wash residue with water; save in vial

Preprocessing:
- SG 2nd degree, 2 poly, windowsize 5 or 6
- EMSC linear (probably needs to involve background spectra to filter them out)

PCA on grains for exploration

PLS regression of spectra on depth

repeat on different grain averages (1, 3, 5, 10)

average all spectra and stack plots of different depth

Pollen extracted from sediment samples 
[@kampsAgentSeparationDissolved1985] SPT patent
[@nakagawaDensemediaSeparationMore1998] 
[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4


# Notes
extracted fossil pollen from sediment.
ca 5 common taxa: Alnus, Picea, Pinus,

## Cores
Dalmuttlado:
- pine
- holocene

Younger Dryas [@brauerHighResolutionSediment1999]:
- 12 samples laminated sediment
- from  11400 to 13120
- min 500 grains
- pollen conc highest in 11000-11600 and 12700-12900
- Pinus betula most common (ca 150 grains counted)
- salix (ca 25 counted)
- quercus throughout in low conc (<1%)
- graminoids/cyperaceae/artemisia common

Tiefer See [@theuerkaufEffectsChangesLand2015]:
- 1870-2010
- laminated sediments
- Pinus pollen

Research Questions:
1. chemical variation between taxa in the same cores.
2. chemical variation between taxa between cores 
3. chemical variation of pollen with depths/age

## Labwork Report

did µFTIR on fresh and fossil pollen using Hyperion-FPA

This is very sample dependent.
organic rich cores are problematic (requires a better extraction protocol), pollen density is another.
Dalmutladdo was a very good core for this: low in organic and relatively rich in pollen.
BSi increased with depth, highest in oldest samples (extracted blanks).
Still, some depth were more time intensive (130/150), low amount of pollen .
I managed about 2-3 core samples per day, the fresh samples can be done in a fraction of the time, all done within 1 half day.

9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (Table \@ref(tab:pollencounts), Table \@ref(tab:fresh))

possibilities for Alnus and Betula, even though scattering visible in spectra 

```{r pollencounts, echo=FALSE}
dalmutladdo = data.frame(ID = c("DAL_08","DAL_20","DAL_40","DAL_60","DAL_90", "DAL_110","DAL_130","DAL_140","DAL_150"),
                         pinus_pollen = c(23,49,37,47,36,32,39,17,7),
                         alnus_pollen = c(0,7,0,0,0,0,0,0,0),
                         betula_pollen = c(0,15,0,0,0,0,0,0,0))
knitr::kable(dalmutladdo, format="latex", booktabs=TRUE,
             caption = "Pollen counts")
```

```{r fresh, echo=FALSE}
fresh_pollen = data.frame(ID = c("BGO_PS7", "BGO_PS8", "INN_PS13", "INN_PS19A","NMBU_Pin_syl_14E1", "NMBU_Pin_syl_14H1", "NMBU_Aln_glu_14E1", "NMBU_Aln_glu_14J1", "NMBU_Aln_inc_14B1", "NMBU_Aln_inc_14G1", "NMBU_Bet_pub_14A1", "NMBU_Bet_pub_14B1", "NMBU_Cor_ave_14B1", "NMBU_Cor_ave_14C1"),
                          extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
knitr::kable(fresh_pollen, format="latex", booktabs=TRUE,
             caption = "Fresh pollen counts")
```

9 fossil samples from 1 core

2 trees of Pinus at each location

Research questions we could answer with this set:

- Differences between fresh and fossil/near fossil grains of Pinus
- chemical changes in Pinus pollen with age/depth
(- chemical differences between Pinus/Alnus/Betula)

extensions to this set:
- add more samples from Dalmuttladdo highest and lowest depth (maybe labwork)
- add cores (e.g. MFM, Tiefer See or Italian cores), Festi core was quite organic rich
- add more grains from already measured samples (to hit min 25 or 50)


# References
